---
title: 'Homework 2'
author: "Jacob Thielemier"
date: "7 February 2024"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE
                      )
```

### Question 1

4.17 is

$$
p_k(x) = \frac{\pi_k\frac{1}{\sqrt{2\pi\sigma}} \exp(-\frac{1}{2\sigma^2}(x - \mu_k)^2)}
              {\sum_{l=1}^k \pi_l\frac{1}{\sqrt{2\pi\sigma}} \exp(-\frac{1}{2\sigma^2}(x - \mu_l)^2)}
$$

and the discriminant function is

$$
\delta_k(x) = x.\frac{\mu_k}{\sigma^2} - \frac{\mu_k^2}{2\sigma_2} + \log(\pi_k)
$$

$\sigma^2$ is constant

$$
p_k(x) = \frac{\pi_k \exp\left(-\frac{1}{2\sigma^2}(x - \mu_k)^2\right)}
              {\sum_{l=1}^k \pi_l \exp\left(-\frac{1}{2\sigma^2}(x - \mu_l)^2\right)}
$$

Maximizing $p_k(x)$ also maximizes $p_k(X)$, so maximize $\log(p_K(X))$

$$
\log(p_k(x)) = \log(\pi_k) - \frac{1}{2\sigma^2}(x - \mu_k)^2 -
              \log\left(\sum_{l=1}^k \pi_l \exp\left(-\frac{1}{2\sigma^2}(x - \mu_l)^2\right)\right)
$$

Maximize over $k$, the last term does not vary with $k$ so it can be ignored. Now maximize

$$
\begin{aligned}
f &= \log(\pi_k) - \frac{1}{2\sigma^2} (x^2 - 2x\mu_k + \mu_k^2) \\
  &= \log(\pi_k) - \frac{x^2}{2\sigma^2} + \frac{x\mu_k}{\sigma^2} - \frac{\mu_k^2}{2\sigma^2} \\
\end{aligned}
$$

$\frac{x^2}{2\sigma^2}$ is independent of $k$

$$
\log(\pi_k) + \frac{x\mu_k}{\sigma^2} - \frac{\mu_k^2}{2\sigma^2}
$$

### Question 2

Same as last question

$$
p_k(x) = \frac{\pi_k\frac{1}{\sqrt{2\pi\sigma_k}} \exp(-\frac{1}{2\sigma_k^2}(x - \mu_k)^2)}
              {\sum_{l=1}^k \pi_l\frac{1}{\sqrt{2\pi\sigma_l}} \exp(-\frac{1}{2\sigma_l^2}(x - \mu_l)^2)}
$$

Derive the Bayes classifier, without assuming $\sigma_1^2 = ... = \sigma_K^2$

Maximizing $p_k(x)$ also maximizes $p_k(X)$, so maximize $\log(p_K(X))$

$$
\log(p_k(x)) = \log(\pi_k) + \log\left(\frac{1}{\sqrt{2\pi\sigma_k}}\right) - \frac{1}{2\sigma_k^2}(x - \mu_k)^2 -
              \log\left(\sum_{l=1}^k \frac{1}{\sqrt{2\pi\sigma_l}} \pi_l \exp\left(-\frac{1}{2\sigma_l^2}(x - \mu_l)^2\right)\right)
$$

Maximizing over $k$, and since the last term does not vary with $k$ it can be ignored. So maximize

$$
\begin{aligned}
f &= \log(\pi_k) + \log\left(\frac{1}{\sqrt{2\pi\sigma_k}}\right) - \frac{1}{2\sigma_k^2}(x - \mu_k)^2 \\
  &= \log(\pi_k) + \log\left(\frac{1}{\sqrt{2\pi\sigma_k}}\right) - \frac{x^2}{2\sigma_k^2} + \frac{x\mu_k}{\sigma_k^2} - \frac{\mu_k^2}{2\sigma_k^2}  \\
\end{aligned}
$$

Now $\frac{x^2}{2\sigma_k^2}$ is not independent of $k$, so retain the term with $x^2$, therefore $f$, the Bayes’ classifier, is a quadratic function of $x$.

### Question 3

#### (a)

```{r, message = FALSE, warning = FALSE}
library(MASS)
library(class)
library(tidyverse)
library(corrplot)
library(ISLR2)
library(e1071)
```

```{r}
summary(Weekly)
corrplot(cor(Weekly[, -9]), type = "lower", diag = FALSE, method = "ellipse")
```

Volume is strongly positively correlated with Year. Other correlations are week, but Lag1 is negatively correlated with Lag2 but positively correlated with Lag3.

#### (b)

```{r}
fit <- glm(
  Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume,
  data = Weekly,
  family = binomial
)
summary(fit)
```

Lag2 is significant.

#### (c)

```{r}
contrasts(Weekly$Direction)
pred <- predict(fit, type = "response") > 0.5
(t <- table(ifelse(pred, "Up (pred)", "Down (pred)"), Weekly$Direction))
sum(diag(t)) / sum(t)
```

The overall fraction of correct predictions is 0.56. Although logistic regression correctly predicts upwards movements well, it incorrectly predicts most downwards movements as up.

#### (d)

```{r}
train <- Weekly$Year < 2009

fit <- glm(Direction ~ Lag2, data = Weekly[train, ], family = binomial)
pred <- predict(fit, Weekly[!train, ], type = "response") > 0.5
(t <- table(ifelse(pred, "Up (pred)", "Down (pred)"), Weekly[!train, ]$Direction))
sum(diag(t)) / sum(t)
```

#### (e)

```{r}
fit <- lda(Direction ~ Lag2, data = Weekly[train, ])
pred <- predict(fit, Weekly[!train, ], type = "response")$class
(t <- table(pred, Weekly[!train, ]$Direction))
sum(diag(t)) / sum(t)
```

#### (f)

```{r}
fit <- qda(Direction ~ Lag2, data = Weekly[train, ])
pred <- predict(fit, Weekly[!train, ], type = "response")$class
(t <- table(pred, Weekly[!train, ]$Direction))
sum(diag(t)) / sum(t)
```

#### (g)

```{r}
fit <- knn(
  Weekly[train, "Lag2", drop = FALSE],
  Weekly[!train, "Lag2", drop = FALSE],
  Weekly$Direction[train]
)
(t <- table(fit, Weekly[!train, ]$Direction))
sum(diag(t)) / sum(t)
```

#### (h)

```{r}
fit <- naiveBayes(Direction ~ Lag2, data = Smarket, subset = train)
pred <- predict(fit, Weekly[!train, ], type = "class")
(t <- table(pred, Weekly[!train, ]$Direction))
sum(diag(t)) / sum(t)
```

#### (i)

Logistic regression and LDA are the best performing.

#### (j)

```{r}
fit <- glm(Direction ~ Lag1, data = Weekly[train, ], family = binomial)
pred <- predict(fit, Weekly[!train, ], type = "response") > 0.5
mean(ifelse(pred, "Up", "Down") == Weekly[!train, ]$Direction)

fit <- glm(Direction ~ Lag3, data = Weekly[train, ], family = binomial)
pred <- predict(fit, Weekly[!train, ], type = "response") > 0.5
mean(ifelse(pred, "Up", "Down") == Weekly[!train, ]$Direction)

fit <- glm(Direction ~Lag4, data = Weekly[train, ], family = binomial)
pred <- predict(fit, Weekly[!train, ], type = "response") > 0.5
mean(ifelse(pred, "Up", "Down") == Weekly[!train, ]$Direction)

fit <- glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4, data = Weekly[train, ], family = binomial)
pred <- predict(fit, Weekly[!train, ], type = "response") > 0.5
mean(ifelse(pred, "Up", "Down") == Weekly[!train, ]$Direction)

fit <- glm(Direction ~ Lag1 * Lag2 * Lag3 * Lag4, data = Weekly[train, ], family = binomial)
pred <- predict(fit, Weekly[!train, ], type = "response") > 0.5
mean(ifelse(pred, "Up", "Down") == Weekly[!train, ]$Direction)

fit <- lda(Direction ~ Lag1 + Lag2 + Lag3 + Lag4,data = Weekly[train, ])
pred <- predict(fit, Weekly[!train, ], type = "response")$class
mean(pred == Weekly[!train, ]$Direction)

fit <- qda(Direction ~ Lag1 + Lag2 + Lag3 + Lag4, data = Weekly[train, ])
pred <- predict(fit, Weekly[!train, ], type = "response")$class
mean(pred == Weekly[!train, ]$Direction)

fit <- naiveBayes(Direction ~ Lag1 + Lag2 + Lag3 + Lag4, data = Weekly[train, ])
pred <- predict(fit, Weekly[!train, ], type = "class")
mean(pred == Weekly[!train, ]$Direction)

set.seed(1)
res <- sapply(1:30, function(k) {
  fit <- knn(
    Weekly[train, 2:4, drop = FALSE],
    Weekly[!train, 2:4, drop = FALSE],
    Weekly$Direction[train],
    k = k
  )
  mean(fit == Weekly[!train, ]$Direction)
})
plot(1:30, res, type = "o", xlab = "k", ylab = "Fraction correct")
(k <- which.max(res))

fit <- knn(
  Weekly[train, 2:4, drop = FALSE],
  Weekly[!train, 2:4, drop = FALSE],
  Weekly$Direction[train],
  k = k
)
table(fit, Weekly[!train, ]$Direction)
mean(fit == Weekly[!train, ]$Direction)
```

KNN using the first 3 Lag variables performs marginally better than logistic regression with `Lag2` if we tune $k$ to be $k = 26$.

### Question 4

#### (a)

```{r}
x <- cbind(Auto[, -1], data.frame("mpg01" = Auto$mpg > median(Auto$mpg)))
```

#### (b)

```{r}
par(mfrow = c(2, 4))
for (i in 1:7) hist(x[, i], breaks = 20, main = colnames(x)[i])

par(mfrow = c(2, 4))
for (i in 1:7) boxplot(x[, i] ~ x$mpg01, main = colnames(x)[i])

pairs(x[, 1:7])
```

Most variables show an association with `mpg01` category, and several variables are colinear.

#### (c)

```{r}
set.seed(1)
train <- sample(seq_len(nrow(x)), nrow(x) * 2/3)
```

#### (d)

```{r}
sort(sapply(1:7, function(i) {
  setNames(abs(t.test(x[, i] ~ x$mpg01)$statistic), colnames(x)[i])
}))

fit <- lda(mpg01 ~ cylinders + weight + displacement, data = x[train, ])
pred <- predict(fit, x[-train, ], type = "response")$class
mean(pred != x[-train, ]$mpg01)
```

#### (e)

```{r}
fit <- qda(mpg01 ~ cylinders + weight + displacement, data = x[train, ])
pred <- predict(fit, x[-train, ], type = "response")$class
mean(pred != x[-train, ]$mpg01)
```

#### (f)

```{r}
fit <- glm(mpg01 ~ cylinders + weight + displacement, data = x[train, ], family = binomial)
pred <- predict(fit, x[-train, ], type = "response") > 0.5
mean(pred != x[-train, ]$mpg01)
```

#### (g)

```{r}
fit <- naiveBayes(mpg01 ~ cylinders + weight + displacement, data = x[train, ])
pred <- predict(fit, x[-train, ], type = "class")
mean(pred != x[-train, ]$mpg01)
```

#### (h)

```{r}
res <- sapply(1:50, function(k) {
  fit <- knn(x[train, c(1, 4, 2)], x[-train, c(1, 4, 2)], x$mpg01[train], k = k)
  mean(fit != x[-train, ]$mpg01)
})
names(res) <- 1:50
plot(res, type = "o")
res[which.min(res)]
```

For the models tested here, $k = 32$ appears to perform best. QDA has a lower error rate overall, performing slightly better than LDA.
