---
title: 'Homework 3'
author: "Jacob Thielemier"
date: "7 April 2024"
output: pdf_document
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	error = FALSE,
	message = FALSE,
	warning = FALSE,
	results = 'hide'
)
```

## Question 1

#### Part (a and b)

```{r}
library(ggplot2)
library(knitr)
library(dplyr)
library(kableExtra)
library(tidyr)
library(readr)
library(broom)

dat <- read.csv("C:\\Users\\JThie\\OneDrive\\Desktop\\Spring 24\\Linear Models\\Homework\\HW3\\framingham.csv")
```

```{r}
colnames(dat)
print(colnames(dat))
```

-   The names of the variables in the data set are:
    -   male
    -   age
    -   education
    -   currentSmoker
    -   cigsPerDay
    -   BPMeds
    -   prevalentStroke
    -   prevalentHyp
    -   diabetes
    -   totChol
    -   sysBP
    -   diaBP
    -   BMI
    -   heartRate
    -   glucose
    -   TenYearCHD

#### Part (c)

```{r}
nrow(dat)
```

-   There are 4240 observations.

#### Part (d)

-   Proposed model: sysBP \~ totChol + male + age + diabetes + diaBP + BMI + currentSmoker + cigsPerDay + glucose

-   The following are possible confounders that are associated with sysBP and totChol:

    -   male
    -   age
    -   diabetes
    -   diaBP
    -   BMI

-   The following are precision variables that will help our estimates:

    -   currentSmoker
    -   cigsPerDay
    -   glucose

## Question 2

#### Part (a)

```{r}
hist(dat$age, main = "Historgram of Age")
hist(dat$cigsPerDay, main = "Histogram of Cigarettes Per Day")
hist(dat$totChol, breaks=50, main = "Histogram of Total Cholesterol")
hist(dat$sysBP, breaks=50, main = "Histogram of Systolic BP")
hist(dat$diaBP, breaks=50, main = "Histogram of Diastolic BP")
hist(dat$BMI, breaks=50, main = "Histogram of BMI")
hist(dat$heartRate, breaks=30, main = "Histogram of Heartrate")
hist(dat$glucose, breaks=25, main = "Histogram of Glucose")
```

#### Part (b)

-   **Age:** looks fine with a small amount being very young.
-   **Cigarettes per day:** Most people that smoke have around 20 cigarettes per day.
-   **Total cholesterol:** It is bell shaped curved, but seems high since typically 200 is ideal
-   **Systolic BP and Diastolic BP:** Both look typical.
-   **BMI:** Typically want to see between 18 and 25. This data seems a little higher, but not concerning.
-   **Heart Rate:** This looks average
-   **Glucose:** The typical range is 70-100 so most data falls in the range. However, the data over 150 does raise concerns.

## Question 3

```{r results='asis'}
categorical_vars <- c('male', 'education', 'currentSmoker', 'BPMeds', 'diabetes')
combined_freq_table <- data.frame() 

for (var in categorical_vars) {
  freq_table <- dat %>%
    count(!!sym(var), name = "Count") %>%
    mutate(
      Percentage = Count / sum(Count) * 100,
      Variable = var, 
      Value = as.character(!!sym(var)) 
    ) %>%
    select(Variable, Value, Count, Percentage) 
  
  if (var == categorical_vars[1]) {
    combined_freq_table <- freq_table
  } else {
    combined_freq_table <- rbind(combined_freq_table, freq_table) 
  }
}

kable(combined_freq_table,
      caption = "Combined Frequency Distribution",
      booktabs = TRUE)
```

## Question 4

```{r}
variables <- c('BMI', 'heartRate', 'glucose', 'totChol', 'sysBP', 'diaBP')

for (var in variables) {
  print(
    ggplot(data = dat, aes_string(x = 'factor(TenYearCHD)', y = var)) +
      geom_boxplot(aes(fill = factor(TenYearCHD))) + 
      scale_fill_manual(values = c("darkgreen", "darkorange"), name = "Ten Year CHD Risk", labels = c("Low", "High")) +
      labs(title = paste('Boxplot of', var, 'by TenYearCHD'),
           x = 'Ten Year CHD Risk',
           y = var) +
      theme_light()
  )
}
```

## Question 5

#### Part (a)

```{r results='asis'}
continuous_vars <- c('age', 'cigsPerDay', 'BMI', 'heartRate', 'glucose', 'totChol', 'sysBP', 'diaBP')  
summary_df <- data.frame(Variable = character(), Mean = numeric(), SD = numeric())

for (var in continuous_vars) {
  mean_val <- mean(dat[[var]], na.rm = TRUE)
  sd_val <- sd(dat[[var]], na.rm = TRUE)
  
  mean_val_rounded <- round(mean_val, 3)
  sd_val_rounded <- round(sd_val, 3)
  
   summary_df <- rbind(summary_df, data.frame(Variable = var, Mean = mean_val, SD = sd_val))
}

kable(summary_df, caption = "Summary Statistics for Continuous Variables", booktabs = TRUE)
```

#### Part (b)

```{r results='asis'}
categorical_vars <- c('male', 'education', 'currentSmoker', 'BPMeds', 'prevalentStroke', 'prevalentHyp', 'diabetes', 'TenYearCHD') 

categorical_summary <- lapply(categorical_vars, function(var) {
  dat %>%
    count(.data[[var]], name = "Count") %>%
    mutate(
      Percentage = round(Count / sum(Count) * 100, 3),
      Variable = paste(var, .data[[var]], sep = " ") 
    ) %>%
    select(Variable, Count, Percentage)
}) %>%
  bind_rows()

kable(categorical_summary, caption = "Frequency Distribution for Categorical Variables") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Question 6

```{r}
na_count_by_variable <- colSums(is.na(dat))
print(na_count_by_variable)

data_clean <- na.omit(dat)

print(nrow(dat))
print(nrow(data_clean))
```

-   Original dataset rows: 4240
-   Rows after removal: 3658

## Question 7

```{r results='asis'}
ggplot(data_clean) + 
  geom_point(aes(x = sysBP, y = totChol, alpha = 0.5)) + 
  labs(x = "Systolic Blood Pressure", y = "Total Cholesterol", title = "Scatterplot of Total Cholesterol by Systolic Blood Pressure") +
  theme_bw()
```

## Question 8

```{r results='asis'}
ggplot(data_clean) + 
  geom_point(aes(x = sysBP, y = totChol, color = factor(male), alpha = 0.5)) + 
  labs(x = "Systolic Blood Pressure", y = "Total Cholesterol", title = "Scatterplot Stratified by Gender", color = "Gender") +
  scale_color_manual(labels = c("Female", "Male"), values = c("darkorange", "darkgreen")) +
  theme_bw()
```

## Question 9

```{r results='asis'}
ggplot(data_clean) + 
  geom_point(aes(x = sysBP, y = totChol, color = factor(currentSmoker), alpha = 0.5)) + 
  labs(x = "Systolic Blood Pressure", y = "Total Cholesterol", title = "Scatterplot Stratified by Current Smoking Status", color = "Current Smoker") +
  scale_color_manual(labels = c("Non-Smoker", "Smoker"), values = c("darkorange", "darkgreen")) +
  theme_bw()
```

## Question 10

```{r results='asis'}
ggplot(data_clean) + 
  geom_point(aes(x = sysBP, y = totChol, color = factor(diabetes), alpha = 0.5)) + 
  labs(x = "Systolic Blood Pressure", y = "Total Cholesterol", title = "Scatterplot Stratified by Current Diabetes Status", color = "Diabetes") +
  scale_color_manual(labels = c("No Diabetes", "Diabetes"), values = c("darkorange", "darkgreen")) +
  theme_bw()
```

## Question 11

-   After conducting the exploratory analysis we can see that the data set is fairly average. The data includes smokers that smoke about 20 cigarettes a day and have a slightly above average **BMI**. Most of the patients also have an elevated amount of **total cholesterol**. This is typical since we are studying the affect of **total cholesterol** on **systolic BP**. There are no outliers of concern at this time.

## Question 12

```{r}
fit1 <- lm(sysBP ~ totChol, data = data_clean)
summary(fit1)
```

```{r results='asis'}
results <- data.frame(
  Term = c("Intercept", "Total Cholesterol"),
  Estimate = c(107.9, 0.103),
  `Std. Error` = c(1.798, 0.007465),
  `P-value` = c("< 0.001", "< 0.001")
)

kable(results, digits = c(NA, 3, 3, NA))
```

## Question 13

- When comparing the the simple model with the new model we see that the p-value still shows that **total cholesterol** is significant. However, there is a decrease in the estimate of **total cholesterol** that means the introduction of **BMI** and **current smoker status** to the model improves the model.

```{r}
fit2 <- lm(sysBP ~ totChol + BMI + currentSmoker, data = data_clean)
summary(fit2)
```

```{r results='asis'} 
results2 <- data.frame(
  Term = c("Intercept", "Total Cholesterol", "BMI", "Current Smoker"),
  Estimate = c(72.76883, 0.08483, 1.59197, -3.26698),
  `Std. Error` = c(2.57945, 0.00711, 0.07880, 0.63898),
  `t value` = c(28.211, 11.931, 20.203, -5.113),
  `P value` = c("< 0.001", "< 0.001", "< 0.001", "< 0.001")
)

knitr::kable(
  results2, 
  digits = c(NA, 3, 3, 3, NA),
  caption = "Linear Regression Model Coefficients"
)
```



## Question 14

```{r}
fit3 <- lm(sysBP ~ totChol + male + age + diabetes + diaBP + BMI + currentSmoker + cigsPerDay + glucose, data = data_clean)
summary(fit3)

fit4 <- lm(sysBP ~ totChol + male + age + diaBP + BMI + glucose, data = data_clean)
summary(fit4)
```

```{r results='asis'} 
results3 <- data.frame(
  Term = c("Intercept", "Total Cholesterol", "Male", "Age", "Diabetes", "DiaBP", "BMI", "Current Smoker", "Cig Per Day", "Glucose"),
  Estimate = c(-17.065382, 0.010726, -3.888875, 0.590871, 1.439938, 1.347187, 0.113468, 0.351229, 0.039500, 0.050553),
  Std_Error = c(2.182805, 0.004749, 0.436348, 0.025435, 1.575867, 0.018663, 0.054737, 0.649970, 0.027864, 0.010812),
  P_value = c("< 0.001", "0.0240", "< 0.001", "< 0.001", "0.3609", "< 0.001", "0.0382", "0.5890", "0.1564", "< 0.001")
)

knitr::kable(
  results3, 
  digits = c(NA, 3, 3, 3),
  caption = "Linear Regression Model Coefficients"
)
```

## Question 15

- The goodness of fit test shows that when comparing my model with the simple model, my model provides as significantly better fit for predicting **systolic BP**. This can be interpreted by the reduction in RSS. This indicates that my model explains a much larger portion of the variance in **systolic BP**. The significant F-statistic from the summary also suggests that the predictors added to my model improve the overall model.

```{r}
anova(fit1, fit3)
anova(fit4, fit3)
```

```{r results='asis'}
anova_results <- data.frame(
  Model = c("Model 1: sysBP ~ totChol", "Model 2: sysBP ~ totChol + male + age + diabetes + diaBP + BMI + currentSmoker + cigsPerDay + glucose"),
  `Res.Df` = c(3656, 3648),
  `RSS` = c(1697707, 566305),
  `Df` = c("-", 8),
  `Sum of Sq` = c("-", 1131402),
  `F` = c("-", 911.03),
  `Pr(>F)` = c("-", "< 0.001")
)

knitr::kable(
  anova_results,
  caption = "ANOVA Results for Comparing Two Models",
  digits = c(NA, 3, 3, 3, 3, 3, NA),
  align = 'lcccclc',
  col.names = c("Model", "Res.Df", "RSS", "Df", "Sum of Sq", "F", "P-value")
)
```


## Question 16

- The Residuals vs. Fitted plot might show a slight funnel shape. This plot indicates the equal variance, but our data does not seem to be concerning.
- The Q-Q plot is the only one that indicates a linear regression assumption may be violated. The plot shows the the tail ends starting to turn upwards. This could indicate that the normaility assumption is violated.
- No comments on the other two plots as they look normal.

```{r results='asis'}
predicted_values <- predict(fit3)

plot(fit3)

plot(data_clean$male, predicted_values, main = "Predicted Values vs. Male", xlab = "Male", ylab = "Predicted Values")
plot(data_clean$age, predicted_values, main = "Predicted Values vs. Age", xlab = "Age", ylab = "Predicted Values")
plot(data_clean$diabetes, predicted_values, main = "Predicted Values vs. Diabetes", xlab = "Diabetes", ylab = "Predicted Values")
plot(data_clean$diaBP, predicted_values, main = "Predicted Values vs. DiaBP", xlab = "DiaBP", ylab = "Predicted Values")
plot(data_clean$BMI, predicted_values, main = "Predicted Values vs. BMI", xlab = "BMI", ylab = "Predicted Values")
plot(data_clean$currentSmoker, predicted_values, main = "Predicted Values vs. Current Smoker", xlab = "Current Smoker", ylab = "Predicted Values")
plot(data_clean$cigsPerDay, predicted_values, main = "Predicted Values vs. Cigs Per Day", xlab = "Cigs Per Day", ylab = "Predicted Values")
plot(data_clean$glucose, predicted_values, main = "Predicted Values vs. Glucose", xlab = "Glucose", ylab = "Predicted Values")
```

## Question 17

- The proposed model indicates that there is a significant association between cholesterol and blood pressure. The Linear Regression Model Summary table below highlights the reported significance. The lack of statistical significance for **diabetes** (P-value=0.36) and current **smoking status** (P-value=0.589) was alarming. Typically someone would assume these variables would impact a persons total cholesterol levels.

- I compared an additional model of **sysBP ~ totChol + male + age + diaBP + BMI + glucose** to my original model. While I removed some covariates that where not significant in my original model, this did not make the additional model better than my original model in a goodness of fit test. This indicates to me that while **diabetes, currentSmoker, and cigsPerDay** may not be significant to the model, they are precision variables that increase the overall effectiveness of the model.

```{r}
conf_intervals <- confint(fit3)
print(conf_intervals)
```

```{r results='asis'} 
ci_lower <- c(-20.901166375, 0.001428025, -4.644803873, 0.524010229, -2.459314438, 1.311754869, -0.002016025, -0.758123794, -0.031481616, 0.031605025)
ci_upper <- c(-12.21209439, 0.02070883, -2.89912159, 0.62604229, 3.86634314, 1.38648984, 0.21693221, 1.83972137, 0.08022605, 0.07459194)

results4 <- results3

results4$`2.5 %` <- ci_lower
results4$`97.5 %` <- ci_upper

results4 <- data.frame(
  Term = c("Intercept", "Total Cholesterol", "Male", "Age", "Diabetes", "DiaBP", "BMI", "Current Smoker", "Cig Per Day", "Glucose"),
  Estimate = c(-17.065382, 0.010726, -3.888875, 0.590871, 1.439938, 1.347187, 0.113468, 0.351229, 0.039500, 0.050553),
  `Std. Error` = c(2.182805, 0.004749, 0.436348, 0.025435, 1.575867, 0.018663, 0.054737, 0.649970, 0.027864, 0.010812),
  `P-value` = c("< 0.001", "0.0240", "< 0.001", "< 0.001", "0.3609", "< 0.001", "0.0382", "0.5890", "0.1564", "< 0.001"),
  `CI_2.5%` = c(-20.901166375, 0.001428025, -4.644803873, 0.524010229, -2.459314438, 1.311754869, -0.002016025, -0.758123794, -0.031481616, 0.031605025),
  `CI_upper` = c(-12.21209439, 0.02070883, -2.89912159, 0.62604229, 3.86634314, 1.38648984, 0.21693221, 1.83972137, 0.08022605, 0.07459194)
)

knitr::kable(
  results3, 
  digits = c(NA, 3, 3, 3, 3, 3, 3),
  caption = "Linear Regression Model Summary"
)
```


